Epistemic Semantics in Guarded String Models\thanks{Thanks guys.}

### 2021-01-16 tex/pdraft.tex
Random edits everywhere.

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a natural
language. A proposition is a set or class of {\em possible worlds}, which are
often assumed to have the same nature and complexity as the world we occupy
\citep{lewis1986plurality}. But alternatively, one can work with small idealized
models, to illustrate and test ideas. The point of this paper is to extend
standard idealized models to countable sets of worlds, and to constructively and
computably model alternatives for epistemic agents.

@@

However, a coin cannot simultanously show both heads and tails! Currently, the
precondition $h$ of $\atHT$ only says that heads must be showing, and says
nothing about the fact that tails must be face-down, indicated by the formula
$\bar t$. The effect valuations are restricted to only the feasible ones via a
\emph{state formula}, shown, and demonstrated for $\apH$ in \ref{effect}. The
state formula says that the coin is either heads-up or tails-up, but not
both\footnote{Two Booleans constrained by a state formula, is more illustrative
  a single boolean $h$ where tails is just $\bar h$.}. Here juxtaposition
represents conjuction, $+$ is disjunction, and overbar is negation.

With the state formula in hand, $\apH\atTH$ is evidently contradictory, since
the post-condition $h$ of $\apH$ is incompatible with the pre-condition $t$ of
$\atTH$. See Figure~\ref{effectfigure} for more details.

@@

Epi\textsc{Kat} is first and foremost a multi-agent epistemic logic, with
epistemic operators that cannot naively be represented as
hypotheses\footnote{Characterizing this relationship precisely is future work.}.
Aside from formalizing the metaphysical modality seen so far,
Sections~\ref{sec:model} and~\ref{sec:logic} develop the epistemic modality.

@@

Rather than individual guarded strings, elements of a guarded string model for
\textsc{Kat} are sets of guarded strings. In Epi\textsc{Kat}, these elements
have the interpretation of propositions, which are (regular) sets of possible
worlds. In a free guarded string model for \textsc{Kat}, any event can be
adjacent to any atom in a guarded string that is an element of the underlying
set for the algebra. However, this subsumes the valid worlds defined by the
state and effect formulas. \ex{wf1} defines the \emph{well-formed} guarded
strings (valid worlds) determined by an Epi\textsc{Kat} specification. Condition
(i) says that each atom is consistent with the state constraint, and condition
(ii) says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with its effects\footnote{ An alternative is to
  define equations such as $\bar{\phi}=0$ (from the state formula $\phi$) and
  $\apH\,{=}\,h \apH h$ (from the effect formula $h\mathbin{:}h$ for event
  $\apH$), and construct a quotient algebra from the equivalence relation
  generated by these equations. This results in equating sets of guarded strings
  in the free algebra that differ by guarded strings that are ill-formed
  according to the state and effect formulas. In the development in the text, we
  instead use a set of guarded strings that are well-formed according to the
  state and effect formulas as the representative of the equivalence class.}.

@@

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics. The set of possible worlds in the modal frame
determined by an Epi\textsc{Kat} specification is the set of well-formed guarded
strings, and propositions are regular sets of guarded strings, rather than the
power set \citep{montague1975formal,gallin1975intensional}. Certain sets of
well-formed guarded strings have the addional interpretation of event types. An
event-type is something that can ``happen'' in different worlds. For example,
$\apH$ corresponds to the event type $\{H\,\apH\,H\}$, and $\nop$ to the event
type $\{T\,\nop\,T, H\,\nop\,H\}$.

The construction so far defines a set of worlds from an Epi\textsc{Kat} specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a general modal frame
$\tuple{W,R_1,...,R_n,K}$ consisting of a set of worlds, a world-alternative relation
for each agent, and a set $K$ of propositions, where each proposition is a subset of
$W$ \citep{chagrov1997modal}.\footnote{As explained in the next section, $K$ will not
  be the power set of $W$, rather it will consists of the regular subsets of $W$.}

@@

Consider a scenario like the one from Figure~\ref{fig:epik-example}, augmented
with an additional agent Cal. The base world $T\,\bpT\,T\,\cpT\,T$ is one where
the coin is tails, and first Bob looks at tails, and them Cal looks at tails.
The first event $\bpT$ has the alternatives $\bpT$ and $\bpH$ for Amy, and the
second event $\bpT$ has the alternatives $\cpT$ and $\cpH$ for Amy. This results
in four combinations $\bpT\cpT$, $\bpT\cpH$, $\bpH\cpT$, and $\bpH\cpH$, some of
which, like $\bpT\cpH$, are bogus. These contradictions are filtered by pre-
and post-conditions of events in the alternative world, so that the set of
alternatives for Amy in $T\,\bpT\,T\,\cpT\,T$ is $\{T\,\bpT\,T\,\cpT\,T,
H\,\bpH\,H\,\cpH\,H\}$, with two world-alternatives instead of four.

@@

With modalities and propositional negation added, the signature of $n$-agent
Epi\textsc{Kat} is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1,
  ..., \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the propositional semantic values of English sentences.

Consider the examples in \ex{interpretation1}. The first term, $\bullet^*\,h$ uses
the syntactic sugar $\bullet$, defined in Figure \ref{fig:epik-grammar} to be
the disjunction of the primitive events. Since a world is a well-formed sequence
of events, $\bullet^*$ is the set of worlds. Multiplying by the state symbol $h$
in the term $\bullet^* h$ has the effect of conjoining $h$ with the atom at the
end of the world. So $\bullet^* h$ is the set of worlds where the coin ends
heads-up.

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\bullet^*t \wedge \Box_a\bullet^*t + \neg\bullet^*t \wedge \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}

@@

{\em Guarded String Interpretation.} A term $p$ of the logical language is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$, where superscript captures dependence
on an Epi\textsc{Kat} specification.  Figure \ref{fig:interp2} defines the
interpretation.  The interpretation $\sem{ 1
}^{\epikspec}$ of the multiplicative identity 1 is the set of atoms that
satisfy the state constraint $\varphi$.  Where $b$ is a primitive
Boolean, $\sem{ b }^{\epikspec}$ is the set of atoms that satisfy the
state constraint and where $b$ is true.  Where $e$ is a primitive
event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have
the form of $e$ flanked by compatible atoms, as determined by the
event formula $\zeta_e$. The product $pq$ is
interpreted with fusion product raised to sets of guarded strings.
Kleene star is interpreted as the union of exponents ($p^n$ is
the $n$-times product of $p$ with itself, with $p^0 = 1$).
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of $p$ 
under the world-alternative relation $\hat{R}_a$.

@@

Summing up, given an Epi\textsc{Kat} specification $\epikspec$, term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a set of
guarded strings $\sem{p}^{\epikspec}$. Let $K^{\epikspec}$ be the sets that are
interpretations of terms. Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of
Epi\textsc{Kat}, with operations as in Figure \ref{fig:interp2}. This provides a
concrete $n$-agent general modal frame $\tuple{\sem{\bullet^*}^{\epikspec},
  \hat{R}_1,...,\hat{R}_n,K^{\epikspec}}$. The frame consists of a set of
worlds, an epistemic-alternative relation for each agent, and a set of
propositions, each of which is a set of worlds. It is used as a target for
natural-language interpretation in Section~\ref{sec:categorial-grammar}.

@@

The space of worlds is a set of ordinary (as opposed to guarded) strings.  Bit
sequences (sequences of $0$'s and $1$'s) encode atoms, and as before, these alternate
with event symbols to encode a world.  In this construction, $w_3$ of the example is
the string {\tt{1\!0\;ah\;1\!0\;bh\;1\!0}}.

@@

Translating the Epi\textsc{Kat} terms $0$, $1$, $b$, and $e$ is
straightforward: convert the atoms as previously described,
and decorate the events $e$ with their compatible atoms. For example, $a_1$ becomes
an Fst term denoting
$\{01\apH01\}$.

Further, Epi\textsc{Kat}'s union $(+)$ and intersection $(\wedge)$ operators can
be directly translated to Fst's built-in analogous operators ((\texttt{|}) and
(\texttt{\&}). Propositional complement in Epi\textsc{Kat} becomes Fst set
difference (\texttt{-}) from the set of all worlds.

Defining \textsc{Kat} product using Fst's set-lifted string concatenation (denoted by
juxtaposition $\texttt{X}~\texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement \textsc{Kat} product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings (Figure \ref{fstfigure1}). {\tt Wf0}
is the set of ordinary strings with only equal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom\footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.}. This relation is
applied via the composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

@@

Finally, the epistemic alternative relation for each agent, i.e. the
\texttt{agent} relation from Figure~\ref{fig:epik-example} is implemented in Fst.
Combining the agent relation with the effect formula gives a relation on events
decorated with compatible atoms. Then the relation on worlds is constructed
using the closure of the concatenation product operation. The concatenation
product $R~S$ of two relations $R$ and $S$ is the set of pairs of the form
$\pair{x_1x_2}{y_1y_2}$, where $x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.
In Fst, {\tt{R+}} is the closure of relation {\tt R} with respect to this
operation. Figure \ref{relclosure} defines the corresponding operation on sets
of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition ({\tt{.o.}}), and the closure under
  discussion here is the closure of the former rather than the latter.} The
epistemic alternative relation on worlds for an agent is then defined as the
\textsc{Kat} relation concatenation closure {\tt RelKst} of the decorated-event
alternative relation for the agent.

@@

English sentences are mapped to terms in the logical language via a semantically
interpreted multi-modal categorial grammar, consisting of a lexicon of words,
their categorial types, and interpretations in a logical lambda language. The
grammar covers basic statives ({\em it's heads}), {\em that-} and {\em
  whether}-complements, negation, and predicate and sentence
conjunction. Figure \ref{lexiconfigure} gives
the lexicon.\footnote{Category symbols use Lambek/Bar-Hillel notation for
  slashes, so that $(d \backslash t)\slash(d \backslash_D t)$ combines with $d
  \backslash_D t$ on the right to give a value that combines with $d$ on the
  left to give $t$. In the semantics, lambda abstractions with multiple
  parameters are written $\lambda x~y.~e$ rather than $\lambda x.\lambda y.e$.
  $d$ is the category of expletive {\em it}.} The online supplement includes a
version of this grammar that can be run using Barker and Shan's chart parser for
categorial grammar \citep{barker2005parser}. The grammar and semantics are
optimized for a simple fragment of English concerned with clausal
complementation. The agent names {\em Amy} and {\em Bob} contribute the
epistemic alternative relations for those agents, rather than individuals. The
root verb {\em believe} contributes existential modal force. The complementizers
{\em that} and {\em whether} are the heads of their dominating clauses, and
assemble an alternative relation, modal force, and proposition contributed by
the complement. These complementizers introduce the dual via two negations, in
order to express universal modal force.

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are
used to control the derivation---phrases with these top-level slashes
can only combine syntactically as arguments.
%
%For instance the category of {\em
%  heads} is $d \backslash_D t$.  The dummy expletive subject {\em it}
%has category $d$.  However the phrase {\em it heads} of category $t$
%can not be formed, because $\backslash_D$ is not syntactically active
%as a function.  Instead {\em it is heads} is formed with a predicator
%{\em is} of category $(d \backslash t)\slash(d \backslash_D t)$.
%Similarly {\em knows} has a category with the top-level slash
%$\slash\!_M$, and combines to form a sentence as an argument of {\em
%  that} or {\em whether}. These complementizers have a category that
%looks for the category of {\em know} on the left, after combining with
%a complement sentence on the right.
The semantic translations in the third column of Figure \ref{lexiconfigure} use
the logical language, incremented with lambda.  The body of $\lambda
x.\bullet^*h$, which is the semantic lexical entry for {\em heads}, is a
term denoting the set of all worlds where the coin is heads, expressed as the
set of all guarded strings that end with a Boolean valuation where the primitive
proposition $h$ (it's heads) is true.
%There is $\lambda x$ at the
%front because the grammar formalism enforces a strict correspondence
%between syntactic and semantic types. However, the lambda does not
%bind anything, because sentences such as {\em it isn't heads} have an
%expletive subject.
The body of $\lambda p.\lambda R. {\Diamond}_R p$, which is the semantic lexical
entry of {\em believes}, is an term denoting the pre-image of the world-alternative
relation contributed by the subject.  This is not the right semantics for {\em Amy
  believes that it's heads}, because it has an existential modality ${\Diamond_R}p$,
rather than an universal modality ${\Box_R}p$.  This is corrected by the
complementizer {\em that}, which introduces the dual.

Sentences are parsed with a chart parser for categorial grammar. The semantics
for complex phrases are obtained by application of semantic translations,
accompanied by beta reductions that eliminate all lambdas in logical forms for
clauses. In consequence, the semantic term translating a sentence is an
Epi\textsc{Kat} logical term. Such a term designates a set of possible words
(guarded strings). By way of example, \exi{three}{a} is an English sentence with
conjunction and several levels of clausal embedding. Using the grammar and
parser, the sentence is mapped to the term in \exi{three}{b}. \exi{three}{c}
shows a simplified logical form constructed from \exi{three}{b} using syntactic
equivalences. Either term is compiled in an implementation of the finite state
calculus to a finite state machine with 10 states and 23 edges, which accepts a
countably infinite set of worlds.\footnote{ Machine sizes need not be small,
  especially as the cardinality of $\sf{B}$ increases. A certain Epi\textsc{Kat}
  model with fourteen primitive Booleans has the set of worlds representented by
  a finite state machine with 184,794 states and 257,881 edges.} In this way the
methodology ``directly'' represents the set of worlds denoted by \exi{three}{a}.

\newcommand{\HEADS}{\bullet^*h}
\newcommand{\TAILS}{\bullet^*t}

\begin{examples} \label{three}
  \itema Amy knows that it's tails and doesn't believe that it's heads and
  believes that Bob believes that it's heads.
\itemb \(
\begin{array}[t]{ll}
  \neg(\neg \TAILS + \Diamond_a \neg \TAILS) & \wedge \\
  \neg \neg \Diamond_a \neg \HEADS & \wedge \\
  \neg \Diamond_a \neg\neg\Diamond_b \neg \HEADS
\end{array}  
\)
\itemc \(\K_{a}\!\TAILS \; \wedge \, \neg\B_{a}\!\HEADS \; \wedge \; \B_{a}\B_{b}\HEADS\)
\itemd Bob believes that it's tails.
\iteme \(\B_{b} \TAILS \)
\end{examples}

@@

\section{Discussion}
Epi\textsc{Kat} is designed for research in linguistic semantics, for
computational linguistics research on model-theoretically grounded semantics,
and for course modules on intensional formal semantics.
Currently, the grammar covers statives, {\em that-} and {\em
  whether}-complements, negation, and conjunction.
There are straightforward extensions to additional linguistic phenomena, such as
tense and perfective aspect \exi{extension}{a}, and the combination of
metaphysical modality and prospective aspect \exi{extension}{b}.

@@

The model framework is a constructive branching-time framework with metaphysical
and epistemic modalities, which will be applicable in linguistic semantic
research on combinations of tense, metaphysical modality, and epistemic
complementation \citep{thomason1984combinations, abusch1998generalizing,
  condoravdi2002temporal}. Connections with research on temporal constitution of
events remain to be explored
\citep{fernando2004finite,fernando2007observing,carlson2009tense}.

>>>>

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a natural
language. A proposition is a set or class of {\em possible worlds}, which are
often assumed to have the same nature and complexity as the world we occupy
\citep{lewis1986plurality}. But alternatively, one can work with small idealized
models, to illustrate and test ideas. The point of this paper is to extend
idealized models to countable sets of worlds, and to constructively and
computably model alternatives for epistemic agents.

@@

However, a coin cannot simultanously show both heads and tails! Currently, the
precondition $h$ of $\atHT$ only says that heads must be showing, and says
nothing about the fact that tails must be face-down, indicated by the formula
$\bar t$. The effect valuations are restricted to only the feasible ones via a
\emph{state formula}, shown and demonstrated for $\apH$ in \ex{effect}. The
state formula says that the coin is either heads-up or tails-up, but not
both\footnote{Two Booleans constrained by a state formula, is more illustrative than
  a single boolean $h$ where tails is just $\bar h$.}. Here juxtaposition
represents conjuction, $+$ is disjunction, and overbar is negation.

With the state formula in hand, $\apH\atTH$ is contradictory, since
the post-condition $h$ of $\apH$ is incompatible with the pre-condition $t$ of
$\atTH$. See Figure~\ref{effectfigure} for more details.

@@

Epi\textsc{Kat} is first and foremost a multi-agent epistemic logic, with
epistemic operators that cannot naively be represented as
hypotheses.\footnote{Characterizing this relationship precisely is future work.}
Aside from formalizing the metaphysical modality seen so far,
Sections~\ref{sec:model} and~\ref{sec:logic} develop the epistemic modality.

@@

Rather than individual guarded strings, elements of a guarded string model for
\textsc{Kat} are sets of guarded strings. In Epi\textsc{Kat}, these elements
have the interpretation of propositions, which are (regular) sets of possible
worlds. In a free guarded string model for \textsc{Kat}, any event can be
adjacent to any atom in a guarded string that is an element of the underlying
set for the algebra. However, this subsumes the valid worlds defined by the
state and effect formulas. \ex{wf1} defines the \emph{well-formed} guarded
strings (valid worlds) determined by an Epi\textsc{Kat} specification. Condition
(i) says that each atom is consistent with the state constraint, and condition
(ii) says that each constituent token event $\alpha_ie_i\alpha_{i+1}$ is
consistent with its effects.\footnote{ An alternative is to
  define equations such as $\bar{\phi}=0$ (from the state formula $\phi$) and
  $\apH\,{=}\,h \apH h$ (from the effect formula $h\mathbin{:}h$ for event
  $\apH$), and construct a quotient algebra from the equivalence relation
  generated by these equations. This results in equating sets of guarded strings
  in the free algebra that differ by guarded strings that are ill-formed
  according to the state and effect formulas. In the development in the text, we
  instead use a set of guarded strings that are well-formed according to the
  state and effect formulas as the representative of the equivalence class.}

@@

Well-formed guarded strings have the interpretation of worlds in the application
to natural-language semantics. The set of possible worlds in the modal frame
determined by an Epi\textsc{Kat} specification is the set of well-formed guarded
strings, and propositions are sets of guarded strings. Certain sets of
well-formed guarded strings have the addional interpretation of event types. An
event-type is something that can ``happen'' in different worlds. For example,
$\apH$ corresponds to the event type $\{H\,\apH\,H\}$, and $\nop$ to the event
type $\{T\,\nop\,T, H\,\nop\,H\}$.

The construction so far defines a set of worlds from an Epi\textsc{Kat} specification.  Normally
the set is countably infinite, though some choices of effect formulas can result in a
finite set of worlds.  The next step is to define an alternative relation $R_a$ on
worlds for each agent $a$. This will result in a general modal frame
$\tuple{W,\hat{R}_1,...,\hat{R}_n,K}$ consisting of a set of worlds, a world-alternative relation
for each agent, and a set $K$ of propositions, where each proposition is a subset of
$W$ \citep{chagrov1997modal}.\footnote{As explained in the next section, $K$ will not
  be the power set of $W$, rather it will consists of the regular subsets of $W$.}

@@

Consider a scenario like the one from Figure~\ref{fig:epik-example}, augmented with
an additional agent Cal. The base world $T\,\bpT\,T\,\cpT\,T$ is one where the coin
is tails, and first Bob looks at tails, and them Cal looks at tails.  The first event
$\bpT$ has the alternatives $\bpT$ and $\bpH$ for Amy, and the second event $\bpT$
has the alternatives $\cpT$ and $\cpH$ for Amy. This results in four combinations
$\bpT,\cpT$, $\bpT,\cpH$, $\bpH,\cpT$, and $\bpH,\cpH$.  These are filtered by pre-
and post-conditions of events in the alternative world, so that the set of
alternatives for Amy in $T\,\bpT\,T\,\cpT\,T$ is $\{T\,\bpT\,T\,\cpT\,T,
H\,\bpH\,H\,\cpH\,H\}$, with two world-alternatives instead of four.

@@

With modalities and propositional negation added, the signature of $n$-agent
epistemic \textsc{Kat} is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1, ...,
  \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language are used to represent the
propositional semantic values of English sentences.  \ex{interpretation1} gives
examples. The abbreviation $\bullet$ as defined in Figure \ref{fig:epik-grammar} is
the disjunction of the primitive events. Since a world is a well-formed sequence of
events, $\bullet^*$ is the set of worlds. Multiplying by the state symbol $h$ in the
term $\bullet^* h$ has the effect of conjoining $h$ with the atom at the end of the
world. So $\bullet^* h$ is the set of worlds where the coin ends heads-up.

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\bullet^*t \wedge \Box_a\bullet^*t + (\neg\bullet^*t) \wedge \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}

@@

{\em Guarded String Interpretation.} A term $p$ of the logical language is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$, where superscript captures dependence
on an Epi\textsc{Kat} specification.  Figure \ref{fig:interp2} defines the
interpretation.  The interpretation $\sem{ 1
}^{\epikspec}$ of the multiplicative identity 1 is the set of atoms that
satisfy the state constraint $\varphi$.  Where $b$ is a primitive
Boolean, $\sem{ b }^{\epikspec}$ is the set of atoms that satisfy the
state constraint and where $b$ is true.  Where $e$ is a primitive
event, $\sem{ e }^{\epikspec}$ is the set of guarded strings that have
the form of $e$ flanked by compatible atoms, as determined by the
effect formula $\zeta_e$. The product $pq$ is
interpreted with fusion product raised to sets of guarded strings.
Kleene star is interpreted as the union of exponents ($p^n$ is
the $n$-times product of $p$ with itself, with $p^0 = 1$).
Propositional complement is complement relative to the set of worlds.
The epistemic formula $\Diamond_a p$ is interpreted with Kripke
semantics for epistemic modality, as the pre-image of $p$ 
under the world-alternative relation $\hat{R}_a$.

@@

Summing up, given an Epi\textsc{Kat} specification $\tuple{\epikspec}$, term $p$ (as
defined syntactically in Figure \ref{termfigure}) is interpreted as a set of guarded
strings $\sem{p}^{\epikspec}$. Let $K^{\epikspec}$ be the sets that are
interpretations of terms. Then
$\tuple{K^{\epikspec},+,\cdot,*,\bar{},0,1,\neg,\Diamond_{a_1},...,\Diamond_{a_n}}$
is a concrete guarded string interpretation for the signature of Epi\textsc{Kat},
with operations as in Figure \ref{fig:interp2}. This provides a concrete $n$-agent
general modal frame $\tuple{\sem{\bullet^*}^{\epikspec},
  \hat{R}_1,...,\hat{R}_n,K^{\epikspec}}$. The frame consists of a set of worlds, an
epistemic-alternative relation for each agent, and a set of propositions, each of
which is a set of worlds. It is used as a target for natural-language interpretation
in Section~\ref{sec:categorial-grammar}.  Since $K^{\epikspec}$ is the set of sets of
worlds that are interpretations of terms, and there are only countably many terms,
not all sets of worlds are propositions in the general modal frame.  Rather,
propositions are regular sets of worlds.

@@

The space of worlds is a set of ordinary (as opposed to guarded) strings.  Bit
sequences (sequences of $0$'s and $1$'s) encode atoms, and as before, these alternate
with event symbols to encode a world.  In this construction, $w_3$ of the example is
the string ``$1 \, 0 \,ah \,1 \,0 \,bh \,1 \,0$''.\footnote{The Fst programming language of the finite
state calculus allows for multicharacter symbols, so that $ah$ is a single symbol in the string.}

@@

Translating the Epi\textsc{Kat} terms $0$, $1$, $b$, and $e$ is straightforward:
convert the atoms as previously described, and decorate the events $e$ with their
compatible atoms. For example, $a_1$ becomes an Fst term denoting $\{01\apH01\}$.
Union $(+)$ and intersection $(\wedge)$ operators are directly translated to Fst's
built-in analogous operators ((\texttt{|}) and (\texttt{\&})). Propositional
complement in Epi\textsc{Kat} becomes Fst set difference (\texttt{-}) from the set of
all worlds.

Defining \textsc{Kat} product using Fst's set-lifted string concatenation (denoted by
juxtaposition $\texttt{X}~\texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement \textsc{Kat} product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings (Figure \ref{fstfigure1}). {\tt Wf0}
is the set of ordinary strings with only equal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom.\footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.} This relation is
applied via the composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

@@

Finally, world alternative relations are constructed in Fst.  Combining the agent
relation with the effect formula gives a relation on events decorated with compatible
atoms. Then the corresponding relation on worlds is constructed using the closure of
the relational concatenation product operation. The concatenation product $R~S$ of
two relations $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$,
where $x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the
closure of relation {\tt R} with respect to this operation. Figure \ref{relclosure}
defines the corresponding operation on sets of guarded strings as encoded in
Fst.\footnote{Relation concatenation in Fst differs from relation composition
  ({\tt{.o.}}), and the closure under discussion here is the closure of the former
  rather than the latter.} The epistemic alternative relation on worlds for an agent
is then defined as the \textsc{Kat} relation concatenation closure {\tt RelKst} of
the decorated-event alternative relation for the agent.

@@

English sentences are mapped to terms in the logical language via a semantically
interpreted multi-modal categorial grammar, consisting of a lexicon of words, their
categorial types, and interpretations in a logical lambda language. The grammar
covers basic statives ({\em it's heads}), {\em that-} and {\em whether}-complements,
negation, and predicate and sentence conjunction. Figure \ref{lexiconfigure} gives
the lexicon.\footnote{Category symbols use Lambek/Bar-Hillel notation for slashes, so
  that $(d \backslash t)\slash(d \backslash_D t)$ combines with $d \backslash_D t$ on
  the right to give a value that combines with $d$ on the left to give $t$. In the
  semantics, lambda abstractions with multiple parameters are written $\lambda
  x~y.~e$ rather than $\lambda x.\lambda y.e$.  $d$ is the category of expletive {\em
    it}.}  The grammar and semantics are optimized for a simple fragment of English
concerned with clausal complementation. The agent names {\em Amy} and {\em Bob}
contribute the epistemic alternative relations for those agents, rather than
individuals. The root verb {\em believe} contributes existential modal force. The
complementizers {\em that} and {\em whether} are the heads of their dominating
clauses, and assemble an alternative relation, modal force, and proposition
contributed by the complement. These complementizers introduce the dual via two
negations, in order to express universal modal force.

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are
used to control the derivation---phrases with these top-level slashes
can only combine syntactically as arguments.
%
%For instance the category of {\em
%  heads} is $d \backslash_D t$.  The dummy expletive subject {\em it}
%has category $d$.  However the phrase {\em it heads} of category $t$
%can not be formed, because $\backslash_D$ is not syntactically active
%as a function.  Instead {\em it is heads} is formed with a predicator
%{\em is} of category $(d \backslash t)\slash(d \backslash_D t)$.
%Similarly {\em knows} has a category with the top-level slash
%$\slash\!_M$, and combines to form a sentence as an argument of {\em
%  that} or {\em whether}. These complementizers have a category that
%looks for the category of {\em know} on the left, after combining with
%a complement sentence on the right.
The semantic translations in the third column of Figure \ref{lexiconfigure} use
the logical language, incremented with lambda.  The body of $\lambda
x.\bullet^*\!h$, which is the semantic lexical entry for {\em heads}, is a
term denoting the set of all worlds where the coin is heads, expressed as the
set of all guarded strings that end with a Boolean valuation where the primitive
proposition $h$ (it's heads) is true.
%There is $\lambda x$ at the
%front because the grammar formalism enforces a strict correspondence
%between syntactic and semantic types. However, the lambda does not
%bind anything, because sentences such as {\em it isn't heads} have an
%expletive subject.
The body of $\lambda p.\lambda R. {\Diamond}_R p$, which is the semantic lexical
entry of {\em believes}, is an term denoting the pre-image of the world-alternative
relation contributed by the subject.  This is not the right semantics for {\em Amy
  believes that it's heads}, because it has an existential modality ${\Diamond_R}p$,
rather than an universal modality ${\Box_R}p$.  This is corrected by the
complementizer {\em that}, which introduces the dual.

Sentences are parsed with a chart parser for categorial grammar. The semantics
for complex phrases are obtained by application of semantic translations,
accompanied by beta reductions that eliminate all lambdas in logical forms for
clauses. In consequence, the semantic term translating a sentence is an
Epi\textsc{Kat} logical term. Such a term designates a set of possible words
(guarded strings). By way of example, \exi{three}{a} is an English sentence with
conjunction and several levels of clausal embedding. Using the grammar and
parser, the sentence is mapped to the term in \exi{three}{b}. \exi{three}{c}
shows a simplified logical form constructed from \exi{three}{b} using syntactic
equivalences. Either term is compiled in an implementation of the finite state
calculus to a finite state machine with 49 states and 110 edges, which accepts a
countably infinite set of worlds.\footnote{ Machine sizes need not be small,
  especially as the cardinality of $\sf{B}$ increases. A certain Epi\textsc{Kat}
  model with fourteen primitive Booleans has the set of worlds representented by
  a finite state machine with 184,794 states and 257,881 edges.} In this way the
methodology ``directly'' represents the set of worlds denoted by \exi{three}{a}.

\newcommand{\HEADS}{\bullet^*\!h}
\newcommand{\TAILS}{\bullet^*\!t}

\begin{examples} \label{three}
  \itema Amy knows that it's tails and doesn't believe that it's heads and
  believes that Bob believes that it's heads.
\itemb \(
\begin{array}[t]{ll}
  \neg(\neg \TAILS + \Diamond_a \neg \TAILS) & \wedge \\
  \neg \neg \Diamond_a \neg \HEADS & \wedge \\
  \neg \Diamond_a \neg\neg\Diamond_b \neg \HEADS
\end{array}  
\)
\itemc \(\K_{a}\!\TAILS \; \wedge \, \neg\B_{a}\!\HEADS \; \wedge \; \B_{a}\B_{b}\HEADS\)
\itemd Bob believes that it's heads.
\iteme \(\B_{b} \HEADS \)
\end{examples}

@@

\section{Discussion}
Epi\textsc{Kat} is designed for research in linguistic semantics, for
computational linguistic research on model-theoretically grounded semantics,
and for course modules on intensional formal semantics.
Currently, the grammar covers statives, {\em that-} and {\em
  whether}-complements, negation, and conjunction.
There are straightforward extensions to additional linguistic phenomena, such as
tense and perfective aspect \exi{extension}{a}, and the combination of
metaphysical modality and prospective aspect \exi{extension}{b}.

@@

The model framework is a constructive branching-time framework with metaphysical
and epistemic modalities, which will be applicable in linguistic semantic
research on combinations of tense, metaphysical modality, and epistemic
complementation \citep{thomason1984combinations, abusch1998generalizing,
  condoravdi2002temporal}. Connections with research in a finite state framework on temporal constitution of
events remain to be explored
\citep{fernando2004finite,fernando2007observing,carlson2009tense}.
